---
title: "Apply citrus for technical factor normalization"
author: "Briana Mittleman"
date: 2017-02-28
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

In this analysis I will use the scPLS method in the Citrus package to regress technical factor out of the single cell data. I will use filtered molecule counts for the first individual in the dataset.  

```{r load packages}
library("Citrus")
library(reshape2)
library(ggplot2) 
#devtools::install_github("kassambara/easyGgplot2")
library(easyGgplot2)
library(stringr)
library("cowplot")
```


###Input data
I will input the data that has been transpormed with 1 + cpm. The ERCC genes and target genes are in seperate datasets. I subset the data for onlt the first individual.  

```{r load data}
molecules_cpm_ERCC= read.table("../data/molecules-cpm-ercc.txt", header=TRUE, stringsAsFactors = FALSE)
molecules_cpm_ERCC_19101= molecules_cpm_ERCC[ ,grep("19101", colnames(molecules_cpm_ERCC) )]
molecules_cpm= read.table("../data/molecules-cpm.txt", header=TRUE, stringsAsFactors = FALSE)
molecules_cpm_19101= molecules_cpm[, grep("19101", colnames(molecules_cpm) )]
```

Transpose datasets for method  

```{r transpose}
molecules_cpm_ERCC_19101t= molecules_cpm_ERCC_19101 %>% t
molecules_cpm_ERCC_19101t[1:3,1:3]
dim(molecules_cpm_ERCC_19101t)
molecules_cpm_19101t = molecules_cpm_19101 %>% t 
molecules_cpm_19101t[1:3, 1:3]
dim(molecules_cpm_19101t)
```

There are 48 ERCC genes and 13058 target genes.  

###Run scPLS  

I will run the scPLS algorithm using the ERCC genes as control and the ENSG genes as the target genes. Start with K values both equaling 1. I will work on optimization of this once I have the algorithm working.    

```{r scPLS, eval=FALSE}
scPLS_tech <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 1, k2 = 1, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
save(scPLS_tech, file="../data/scPLS_tech.rda")
```

```{r load scPLS}
load("../data/scPLS_tech.rda")
```

###Evaluate model  

```{r}
summary(scPLS_tech)
summary(scPLS_tech$Likelihood)
```
```{r make quantile table }
Var <- scPLS_tech$VarianceSummary
ConProp <- round(Var["ModelConfounding"]/Var["Model"], 3)
StructuredProp <- round(Var["ModelStructure"]/Var["Model"], 3) 
PropTable <- data.frame(ConProp, StructuredProp, 1 - ConProp -
StructuredProp)
colnames(PropTable) <- c("Tech", "Structured", "Residue")
QuantileTable <- QuantileSummary(PropTable, quantiles = seq(0.1,
1, by = 0.1), rankingby = unlist(Var["Sample"]))
head(QuantileTable)
```
```{r}
df <- reshape2::melt(QuantileTable, id.vars = "Quantile") 
colnames(df)[3] <- "Proportion"
colnames(df)[2] <- "Variance"

ff <- ggplot2.barplot(data = df, xName = "Quantile", yName = "Proportion", groupName = "Variance", groupColors = c("#66B2FF", "#FFB266",
"#FFAAD4"), position = position_stack(), backgroundColor = "white", color = "black", xtitle = "Quantile (%)", ytitle = "Proportion", mainTitle = "", removePanelGrid = TRUE, removePanelBorder = TRUE, axisLine = c(0.5, "solid", "black"), ylim = c(0, 1.05), legendPosition = "top", legendTextFont = c(10, "bold", "black"))

ff
```



```{r}
celltypes <- str_split_fixed(row.names(molecules_cpm_19101t), pattern= ".[A-H]", n=3)[,2]

pcs <- prcomp(molecules_cpm_19101t, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "raw", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")


ff1 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2", groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1", ytitle = "PC2", mainTitle = "Raw", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = FALSE, mainTitleFont = c(15, "bold", "black"), xtitleFont = c(10,"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont = c(10, "bold", "black"), yTickLabelFont = c(10, "bold", "black"))

pcs <- prcomp(scPLS_tech$Adjusted, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "rm_ERCC", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")

ff2 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2",
groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1",
ytitle = "PC2", mainTitle = "ERCC technical effect removed", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = TRUE, mainTitleFont = c(10, "bold", "black"), xtitleFont = c(10,
"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont=c(10, "bold", "black"), yTickLabelFont= c(10, "bold", "black"), legendTitle= "Cell type", legendTitleFont= c(10, "bold", "black"), legendTextFont = c(10, "bold", "black") )

```


###Plot output  


```{r}
plot_grid(ff1,ff2)
```


###Try more k values  
I changed the chunk size to 500 to try to run the algorithm faster.  

```{r try k values, eval=FALSE}
scPLS_tech.1.2 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 1, k2 = 2, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
#save(scPLS_tech.1.2, file="../data/scPLS_tech.1.2.rda")
scPLS_tech.2.1 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 2, k2 = 1, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
#save(scPLS_tech.2.1, file="../data/scPLS_tech.2.1.rda")
scPLS_tech.2.2 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 2, k2 = 2, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
#save(scPLS_tech.2.2, file="../data/scPLS_tech.2.2.rda")
scPLS_tech.2.3 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 2, k2 = 3, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
save(scPLS_tech.2.3, file="../data/scPLS_tech.2.3.rda")
scPLS_tech.3.2 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 3, k2 = 2, iter = 100, method = "EM", Chunk = TRUE,chunk.size = 500,  center = TRUE)
save(scPLS_tech.3.2, file="../data/scPLS_tech.3.2.rda")
scPLS_tech.3.3 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 3, k2 = 3, iter = 100, method = "EM", Chunk = TRUE,chunk.size = 500,  center = TRUE)
save(scPLS_tech.3.3, file="../data/scPLS_tech.3.3.rda")

```

```{r load k values}
load("../data/scPLS_tech.1.2.rda")
load("../data/scPLS_tech.2.1.rda")
load("../data/scPLS_tech.2.2.rda")
load("../data/scPLS_tech.2.3.rda")
load("../data/scPLS_tech.3.2.rda")
load("../data/scPLS_tech.3.3.rda")
```

###Plot K value likelihoods  

Create dataframe with likelihoods  

```{r vary k dataframe}
scPLS.1.1= c(1,1,mean(summary(scPLS_tech$Likelihood)))
scPLS.1.2 = c(1,2,mean(summary(scPLS_tech.1.2$Likelihood)))
scPLS.2.1 = c(2,1,mean(summary(scPLS_tech.2.1$Likelihood)))
scPLS.2.2 = c(2,2,mean(summary(scPLS_tech.2.2$Likelihood)))
scPLS.2.3 = c(2,3,mean(summary(scPLS_tech.2.3$Likelihood)))
scPLS.3.2 = c(3,2,mean(summary(scPLS_tech.3.2$Likelihood)))
scPLS.3.3 = c(3,3,mean(summary(scPLS_tech.3.3$Likelihood)))
vary_k = rbind(scPLS.1.1, scPLS.1.2, scPLS.2.1, scPLS.2.2, scPLS.2.3, scPLS.3.2, scPLS.3.2)
colnames(vary_k) = c("k1", "k2", "likelihood")
vary_k =data.frame(vary_k)
```

```{r}
vary_k
ggplot2.scatterplot(data=vary_k, xName = "k1", yName = "k2", mapping=aes(size =("likelihood")),  backgroundColor="white")

```


###Do something to compare likliehood 
compare to previous 1 
(one-other)/one 

run: (1,3), (1,4), (1,5)


```{r}
pcs <- prcomp(scPLS_tech.1.2$Adjusted, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "rm_ERCC", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")

ff3 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2",
groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1",
ytitle = "PC2", mainTitle = "ERCC technical effect removed, k2=2", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = TRUE, mainTitleFont = c(10, "bold", "black"), xtitleFont = c(10,
"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont=c(10, "bold", "black"), yTickLabelFont= c(10, "bold", "black"), legendTitle= "Cell type", legendTitleFont= c(10, "bold", "black"), legendTextFont = c(10, "bold", "black") )

ff3
```


* need k1=1 and k2=1:5  

```{r 1.3-5, eval=FALSE}
scPLS_tech.1.3 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 1, k2 = 3, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
save(scPLS_tech.1.3, file="../data/scPLS_tech.1.3.rda")
scPLS_tech.1.4 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 1, k2 = 4, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
save(scPLS_tech.1.4, file="../data/scPLS_tech.1.4.rda")
scPLS_tech.1.5 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 1, k2 = 5, iter = 100, method = "EM", Chunk = TRUE, chunk.size = 500, center = TRUE)
save(scPLS_tech.1.5, file="../data/scPLS_tech.1.5.rda")
```

```{r load 3-5}
load("../data/scPLS_tech.1.3.rda")
load("../data/scPLS_tech.1.4.rda")
load("../data/scPLS_tech.1.5.rda")
```

```{r extract L}
scPLS.1.1= c(1,1,mean(summary(scPLS_tech$Likelihood)))
scPLS.1.2 = c(1,2,mean(summary(scPLS_tech.1.2$Likelihood)))
scPLS.1.3 = c(1,3,mean(summary(scPLS_tech.1.3$Likelihood)))
scPLS.1.4 = c(1,4,mean(summary(scPLS_tech.1.4$Likelihood)))
scPLS.1.5 = c(1,5,mean(summary(scPLS_tech.1.5$Likelihood)))
scPLS.1.1.5= rbind(scPLS.1.1,scPLS.1.2,scPLS.1.3,scPLS.1.4,scPLS.1.5)
colnames(scPLS.1.1.5)=c("k1", "k2", "Likelihood")
scPLS.1.1.5
```
```{r}
change1=(mean(summary(scPLS_tech.1.2$Likelihood))- mean(summary(scPLS_tech$Likelihood)))/mean(summary(scPLS_tech$Likelihood))

change2=(mean(summary(scPLS_tech.1.3$Likelihood))- mean(summary(scPLS_tech.1.2$Likelihood)))/mean(summary(scPLS_tech.1.2$Likelihood))

change3=(mean(summary(scPLS_tech.1.4$Likelihood))- mean(summary(scPLS_tech.1.3$Likelihood)))/mean(summary(scPLS_tech.1.3$Likelihood))

change4=(mean(summary(scPLS_tech.1.5$Likelihood))- mean(summary(scPLS_tech.1.4$Likelihood)))/mean(summary(scPLS_tech.1.4$Likelihood))

L_change= c(abs(change1), abs(change2), abs(change3), abs(change4))
plot(L_change, ylab="Absolute value change in Likelihood", main="Compare k2 values 1-5")

```

Biggest improvement in likelihood is between k2=1 and k2=2. 

**DON'T USE THIS ANALYSIS FOR MODEL SELECTION, USE THE MIN BIC**


###Compare likelihoods with BIC  
BIC= ln(n)k-2ln(L)  
```{r}
#BIC = ln(number of cells) * (k1 + k2) - 2 *(scPLS output of likelihood)
ncells=201

Bic1= log(ncells) * (2) - (2 * (mean(summary(scPLS_tech$Likelihood))))
Bic2= log(ncells) * (3) - (2 * (mean(summary(scPLS_tech.1.2$Likelihood))))
Bic3= log(ncells) * (4) - (2 * (mean(summary(scPLS_tech.1.3$Likelihood))))
Bic4= log(ncells) * (5) - (2 * (mean(summary(scPLS_tech.1.4$Likelihood))))
Bic5= log(ncells) * (6) - (2 * (mean(summary(scPLS_tech.1.5$Likelihood))))

Bic_likelihood= c(Bic1,Bic2, Bic3, Bic4, Bic5)

plot(Bic_likelihood, ylab="Bayesian information criterion", xlab="k2 value", main="With a constant k1 value, BIC decreases with k2 increase")

which.min(Bic_likelihood)
```

The smallest BIC is achieved when k2=5. 


###Plot models with varying K values  

```{r ff4}
#ff1 is raw
#k2=1 : ff2
#k2=2: ff3



pcs <- prcomp(scPLS_tech.1.3$Adjusted, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "rm_ERCC", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")

ff4 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2",
groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1",
ytitle = "PC2", mainTitle = "k2=3", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = TRUE, mainTitleFont = c(10, "bold", "black"), xtitleFont = c(10,
"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont=c(10, "bold", "black"), yTickLabelFont= c(10, "bold", "black"), legendTitle= "Cell type", legendTitleFont= c(10, "bold", "black"), legendTextFont = c(10, "bold", "black") )


```

```{r ff5}
pcs <- prcomp(scPLS_tech.1.4$Adjusted, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "rm_ERCC", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")

ff5 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2",
groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1",
ytitle = "PC2", mainTitle = "k2=4", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = TRUE, mainTitleFont = c(10, "bold", "black"), xtitleFont = c(10,
"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont=c(10, "bold", "black"), yTickLabelFont= c(10, "bold", "black"), legendTitle= "Cell type", legendTitleFont= c(10, "bold", "black"), legendTextFont = c(10, "bold", "black") )
```

```{r ff6}
pcs <- prcomp(scPLS_tech.1.5$Adjusted, center = TRUE)
PC1 <- pcs$x[, 1]
PC2 <- pcs$x[, 2]
df <- data.frame(PC1, PC2, "rm_ERCC", celltypes) 
colnames(df) <- c("PC1", "PC2", "Method", "Type")

ff6 <- ggplot2.scatterplot(data = df, xName = "PC1", yName = "PC2",
groupName = "Type", size = 5, backgroundColor = "white", groupColors = c("#66B2FF", "#FFB266", "#FFAAD4"), xtitle = "PC1",
ytitle = "PC2", mainTitle = "k2=5", removePanelGrid = TRUE, removePanelBorder = TRUE, setShapeByGroupName = TRUE, showLegend = TRUE, mainTitleFont = c(10, "bold", "black"), xtitleFont = c(10,
"bold", "black"), ytitleFont = c(10, "bold", "black"), xTickLabelFont=c(10, "bold", "black"), yTickLabelFont= c(10, "bold", "black"), legendTitle= "Cell type", legendTitleFont= c(10, "bold", "black"), legendTextFont = c(10, "bold", "black") )
```






```{r}
plot_grid(ff1, ff2, ff3, ff4, ff5, ff6, nrow = 3 )
```



###Run k1=2 values 


```{r eval=FALSE}
scPLS_tech.2.4 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 2, k2 = 4, iter = 100, method = "EM", Chunk = TRUE,chunk.size = 500,  center = TRUE)
save(scPLS_tech.2.4, file="../data/scPLS_tech.2.4.rda")
scPLS_tech.2.5 <- scPLS(molecules_cpm_19101t, molecules_cpm_ERCC_19101t, k1 = 2, k2 = 5, iter = 100, method = "EM", Chunk = TRUE,chunk.size = 500,  center = TRUE)
save(scPLS_tech.2.5, file="../data/scPLS_tech.2.5.rda")
```



```{r}
load("../data/scPLS_tech.2.1.rda")
load("../data/scPLS_tech.2.2.rda")
load("../data/scPLS_tech.2.3.rda")
load("../data/scPLS_tech.2.4.rda")
load("../data/scPLS_tech.2.5.rda")
```


```{r}
Bic2.1= log(ncells) * (3) - (2 * (mean(summary(scPLS_tech.2.1$Likelihood))))
Bic2.2= log(ncells) * (4) - (2 * (mean(summary(scPLS_tech.2.2$Likelihood))))
Bic2.3= log(ncells) * (5) - (2 * (mean(summary(scPLS_tech.2.3$Likelihood))))
Bic2.4= log(ncells) * (6) - (2 * (mean(summary(scPLS_tech.2.4$Likelihood))))
Bic2.5= log(ncells) * (7) - (2 * (mean(summary(scPLS_tech.2.5$Likelihood))))
```


```{r}

x1.1= c(1,1, Bic1/(10^9))
x1.2= c(1,2,Bic2/(10^9))
x1.3= c(1,3,Bic3/(10^9))
x1.4= c(1,4,Bic4/(10^9))
x1.5= c(1,5,Bic5/(10^9))

x2.1= c(2,1, Bic2.1/(10^9))
x2.2= c(2,2,Bic2.2/(10^9))
x2.3= c(2,3,Bic2.3/(10^9))
x2.4= c(2,4,Bic2.4/(10^9))
x2.5= c(2,5,Bic2.5/(10^9))


BIC_m= rbind(x1.1, x1.2,x1.3,x1.4,x1.5, x2.1, x2.2, x2.3, x2.4, x2.5)
colnames(BIC_m)=c("k1", "k2", "BIC_prop")
BIC_df=data.frame(BIC_m)
BIC_df



ggplot(BIC_df, aes(x=k1, y=k2)) + 
  geom_point(aes(size=BIC_prop)) +
  ggtitle("BIC values divided by 10^9 plotted by K values") +
  scale_x_discrete("Number of Biological factors", limits=c("1","2")) +
  ylab(" Number of Technical factors ") 
  


      
```
###Final model selection  
 
Final model: k1=1 and k2=5

```{r}
ff6
```

