---
title: "Recreate Gilad lab single cell"
author: "Brana Mittleman"
date: "January 11, 2017"
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`


I will use this to document my recreating of the single cell data processing from http://jdblischak.github.io/singleCellSeq/analysis/process-samples.html.  

####General information for committing changes to the git hub site: 

``` 
#in R
git add analysis/*.Rmd
git commit -m "message"
```
Run `commit_site()` in R terminal.

```
#in R
 git push origin master
```


####Programs for analysis 
First I need to document the programs I will need to add to midway to complete the processing recreation. 

- umitools v2.1.1
- sickle version 1.33
- Subread version 1.5.0-p1 **on midway**
- samtools version 0.1.18-dev (r982:313) **version 1.2 on midway**
- UMI-tools (version e0ade5d)
- featureCounts version 1.5.0-p1 **i think this is a part of subread**

To use modules on midway:
```
module load [name]
module list #show loaded modules 
module unload [name]

```



####Moving the files from pps to midway 
First I copy the first individual, first plate from spudhead to gateway.  
```
scp 19098.1* pps-gateway
```
Move the files to midway:
```
scp 19098.1* brimittleman@midway.rcc.uchicago.edu:
```
Remove the files from gateway: 
```
rm 19098.1*
```
Move to the midway cluster and move the files to a directory in scratch. 
```
mv 19098.1* scratch-midway/19098.1 
```

Create a code directory for scripts. 
```
mkdir code
```
Copy contents of create-genome.sh from John's github to a script in my own directory using vi. 

```
vi create-genome.sh
```
Use escape then :wq to write out file.  

Run each script seperatly:
```
sbatch --mem=8g --wrap="create-genome.sh genome"
```
This submits the job.To view the job I use this command.
```
squeue -u brimittleman
```
