---
title: "Recreate Gilad lab single cell"
author: "Brana Mittleman"
date: "January 11, 2017"
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`


I will use this to document my recreating of the single cell data processing from http://jdblischak.github.io/singleCellSeq/analysis/process-samples.html.  

####General information for committing changes to the git hub site: 

``` 
#in R
git add analysis/*.Rmd
git commit -m "message"
```
Run `commit_site()` in R terminal.

```
#in R
 git push origin master
```


####Programs for analysis 
First I need to document the programs I will need to add to midway to complete the processing recreation. 

- umitools v2.1.1
- sickle version 1.33
- Subread version 1.5.0-p1
- samtools version 0.1.18-dev (r982:313) 
- UMI-tools (version e0ade5d)
- featureCounts version 1.5.0-p1
