---
title: "Simulate bulk samples from Single cell gene list"
author: "Briana Mittleman"
date: 2017-2-16
output: html_document
---

```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

```{r knitr-opts-chunk, include=FALSE}
```

**Last updated:** `r Sys.Date()`

**Code version:** `r workflowr::extract_commit(".", 1)$sha1`

```{r}
library("biomaRt")
library("data.table")
library("testit")
library("dplyr")
library("edgeR")
library("VennDiagram")
library("ggplot2")
theme_set(theme_bw(base_size = 12))
source("../../singleCellSeq/analysis/functions.R")
```

###Input 
```{r}
anno <- read.table("../../singleCellSeq/data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
#input molecule (pre-filter)
molecules <- read.table("../../singleCellSeq/data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
reads_bulk <- read.table("../../singleCellSeq/data/reads-bulk.txt", header = TRUE, stringsAsFactors = FALSE)
```
Create gene lists:  
```{r}
genes_observed_reads_bulk <- rownames(reads_bulk)[rowSums(reads_bulk) > 0]
genes_observed_molecules <- rownames(molecules)[rowSums(molecules) > 0]
```

###Pool molecule counts  
```{r}
molecules_1ind<- molecules[,1:96]
molecules_pool <- data.frame(apply(molecules_1ind, 1 , sum))
colnames(molecules_pool) <- c("pool_count")
#i need a vector containing each of the genes the number of times they appear in the sum for sampling (pooled_molecules)

molecules_1ind_3r <- molecules[,1:288]
molecules1ind3r_pool <- data.frame(apply(molecules_1ind_3r, 1 , sum))
colnames(molecules1ind3r_pool) <- c("pool_count")

```

```{r}


pooled_molecules <- c()
for (row in rownames(molecules_pool)){
  which_row <- rownames(molecules_pool) == row
  num = rep(row, times= molecules_pool[which_row,])
  pooled_molecules <- c(pooled_molecules, num)
}

```
3 replicates  
```{r}

pooled_molecules_3rep <- c()
for (row in rownames(molecules1ind3r_pool)){
  which_row3 <- rownames(molecules1ind3r_pool) == row
  num = rep(row, times = molecules1ind3r_pool[which_row3,])
  pooled_molecules_3rep <- c(pooled_molecules_3rep, num)
}

```


###Sampling function  
I will make a function that samples the gene vector N times and creates a list of the unique gene names in the sample. It will then see how many genes are in the genes_observed_molecules but not in this "simulated" bulk. (sample no replacement)
```{r}
subsample_func <- function(N){
  #subsample with replacement
  subsample_pool <- sample(pooled_molecules, N, replace=TRUE)
  #make list of unique genes in sample
  unique_subsample_pool <- unique(subsample_pool)
  #list of gene in genes_observed_molecules that arnt in unique list (call it only_sc)
  only_sc <- setdiff(genes_observed_molecules, unique_subsample_pool)
  return(length(only_sc))
}
  
```


```{r}
subsample_func_3 <- function(N){
  #subsample with replacement
  subsample_pool <- sample(pooled_molecules_3rep, N, replace=TRUE)
  #make list of unique genes in sample
  unique_subsample_pool <- unique(subsample_pool)
  #list of gene in genes_observed_molecules that arnt in unique list (call it only_sc)
  only_sc <- setdiff(genes_observed_molecules, unique_subsample_pool)
  return(length(only_sc))
}
```



using a multinomial approach  
```{r}
# not_in_simbulk<- sapply(N_samples, function(x) {
#   subsample_genes <- rmultinom(n = x, size = length(molecules_pool$pool_count),
#             prob = molecules_pool$pool_count)
#   unique_subsample_pool <- rownames(molecules_pool)[rowSums(subsample_genes)>0]
#   #list of gene in genes_observed_molecules that arnt in unique list (call it only_sc)
#   only_sc <- length(setdiff(genes_observed_molecules, unique_subsample_pool))
#   return(only_sc)
# })
```

###Apply function  
I will run my function at a bunch of different N values and plot the returned value against the sample size N.  

```{r}
N_samples <-seq(100000, 100000000, by = 1000000)
not_in_simbulk<- sapply(N_samples, function(x) subsample_func(x))
plot(not_in_simbulk ~ N_samples, main="Simulated bulk from 1 single cell individual 1 replicate", xlab="N sampling", ylab = "# genes in single cell list not in simulated bulk") 
abline(971,0)

not_in_simbulk3r<- sapply(N_samples, function(x) subsample_func_3(x))
plot(not_in_simbulk3r ~ N_samples, main="Simulated bulk from 1 single cell individual 3 replicate", xlab="N sampling", ylab = "# genes in single cell list not in simulated bulk") 
abline(971, 0)
points(31279090, 971, cex=1,  col="blue", pch=16)

```
